---
title: "Project 2 - Stochastic Modelling"
output: pdf_document
date: 'By: Sanne Jamila Razmara Olsen & Einride Brodahl Osland'
---
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,strip.white=TRUE,prompt=FALSE,
                      cache=TRUE, size="scriptsize",fig.width=4, fig.height=3,fig.align = "center")
```

__Problem 1 - Urgent care center__

Assume that $\lambda < \mu$, and letting $X(t)$ denote the number of patients in the UCC ar time $t$. We have $X(0) = 0$.

a)

In order for the UCC to be a M/M/1 queue the arrivals of patients must be a Poisson process, the service time must be identically distributed and follow an exponential distribution. We also need the number of customers to be a birth/death process, and the single server must be defined by the first to arrive being the first to be served. 

For our case we assumed that the patients of the UCC follows a Poisson process with rate $\lambda > 0$ and we assumed that the treatment times of the patients follow independent exponential distributions, which means the same as the service time being identically distributed and follow an exponential distribution. We also have that the patients are waiting to be treated in a queue, where the first one gets treated first, meaning that the single server must be defined by the first to arrive being the first to be served. We see that thus the conditions for the UCC to be an M/M/1 queue are satisfied. 

We can perceive the stochastic process as birth/death process, where $X(t)$ increases by one for a new person entering the queue, or decreasing by one for a finished treatment - time until new person arrives and the time until a person is treated are independently exponentially distributed with birth and death rates $\lambda$ and $\mu$.

We now want to find the average time a patient will spend in the UCC as a function of the birth and death rates $\lambda$ and $\mu$.

By Littles law we have $L = \lambda W$, where $L$ denotes the average number of customers, and $W$ denotes the average time spent by a customer in the system (which we want to find).

We have that $W = \frac{L}{\lambda}$.

We have that as $\lambda < \mu$ that $L$ must be on the form $L = \frac{\lambda}{\mu - \lambda}$. We thus have

$$W = \frac{L}{\lambda} = \frac{1}{\mu-\lambda}$$
Must be the average time spent by a customer in the system. 

b)


```{r,eval=TRUE}
lambda = 5
mu = 0.1*60    #Minutes to hours
n = 50

simulate_UCC <- function(lambda,mu,p){
  U = c(0)  #List containing number of urgent patients for each n
  Q = c(0)  #List containing number of "normal" patients for each n
  X = c(0)  #List containing total number of patients
  
  #Defining entering and leaving times
  
  U_entertime = c()
  N_entertime = c()
  
  U_leavetime = c()
  N_leavetime = c()
  
  for(i in 1:n){
    state = tail(X,1)
    if(state == 0){ #No patients in UVV
      times = c(times, (tail(times,1) + rexp(n=1, rate=lambda))) #Time of entering for next person
      X = c(X, tail(X,1)+1) #Total people increment
      q = runif(1)
      #Checking if urgent or normal
      if(q < p){
        #TRUE -> urgent case
        U = c(U, tail(U,1)+1) #Adding urgent person
        U_entertime = c(U_entertime, tail(times,1)) #Adding time of entering
      }
      else{
        #FALSE -> normal case
        N = c(N, tail(N,1)+1) #Adding normal person
        N_entertime = c(N_entertime, tail(times,1)) #Adding time of entering
      }
    }
    else{
      
    }
  }
}


#to be continued
```

c)

We first want to show that $\{ U(t) : t \ge 0 \}$ satisfies the conditions for an M/M/1 queue. 

We have that urgent patients have prioritized treatment, and thus the queuing system for $U(t)$ is the same as $X(t)$. We also have that the queue is defined by the first to come getting treatment first. We still have a Poisson process for this case, but with a different rate than for the former case ($X(t)$). The treatment time is independent of the shift in queue and is thus still exponentially distributed. We still have one server, and we therefore have an M/M/1 queue. 

We now want to determine the arrival rate. We have that the probability of being an urgent patients is $p$, and thus the arrival rate for the urgent patients is $p \lambda$, and it must therefore be $(p-1) \lambda$ for the normal patients.

We now want to determine the long-run mean number of urgent patients in the UCC s a function of $\lambda, \mu$ and $p$.

The long run mean number of urgent patients is given by:

$$
E[U] = \sum_{i = 0}^\infty i \pi_i
$$

The Kolmogorov forward equations gives us

$$
\lambda_0 \pi_0 = \mu_1 \pi_1
$$
and

$$
\mu_{j+1} \pi_{j+1} = (\lambda_j + \mu_j) \pi_j - \lambda_{j-1}\pi_{j-1}, j\ge 1
$$

By $\pi_j = \theta_j \pi_0$, $\theta_j = \frac{(p \lambda)^j}{\mu^j}$, we get

$$
\pi_j = \frac{(p \lambda)^j}{\mu^j} \pi_0
$$
We also have that

$$
\sum_{j=0}^\infty \pi_j = 1 = \pi_0 \sum^\infty_{j=0}\frac{(p \lambda)^j}{\mu^j} = \frac{\pi_0}{1-\frac{p\lambda}{\mu}}
$$

Such that $\pi_0 = 1-\frac{p\lambda}{\mu}$ and thus we have

$$
\pi_j = \frac{(p \lambda)^j}{\mu^j} \pi_0= \frac{(p \lambda)^j}{\mu^j} (1-\frac{p\lambda}{\mu})
$$
We now find $E[U]$.

$$
E[U] = \sum_{i = 0}^\infty i \pi_i = \sum_{i = 0}^\infty i\frac{(p \lambda)^j}{\mu^j} (1-\frac{p\lambda}{\mu}) = \frac{1}{\frac{\mu}{p \lambda}-1}
$$

Thus the long-run mean number of urgent patients is given by

$$
\frac{1}{\frac{\mu}{p \lambda}-1}
$$
d)

We want to explain in words why $\{N(t) : t \ge 0\}$ does not behave as an M/M/1 queue.

The treatment times are not identically distributed. In the case of only normal patients, the treatment time of urgent patients are added to some of the waiting time for normal patients. 

We now want to determine the long-run mean number of normal patient as a function of $\lambda, \mu$ and $p$.

We have that

$E[N] = E[X] - E[U]$

We have that $E[X] = \frac{1}{\frac{\mu}{\lambda}-1}$ such that

$$
E[N] = \frac{1}{\frac{\mu}{\lambda}-1} -\frac{1}{\frac{\mu}{p \lambda}-1} = \frac{\lambda}{\mu - \lambda}- \frac{p\lambda}{\mu - p\lambda} = \frac{(1-p)\lambda \mu}{(\mu- \lambda)(\mu - p \lambda)}
$$

Such that the long-run mean number of normal patients is given by
$$
\frac{(1-p)\lambda \mu}{(\mu- \lambda)(\mu - p \lambda)}
$$
e)

We want to use Littles law to show that 

1. The expected time in the UCC for an urgent patient as a function of $\lambda, \mu$ and $p$ is given by

$$
W_U = \frac{1}{\mu- p \lambda}
$$

and

2. The expected time in the UCC for a normal patient as a function of $\lambda, \mu$ and $p$ is given by

$$
W_N = \frac{\mu}{(\mu-\lambda)(\mu-p \lambda)}
$$

We have from Littles law that $L = \lambda W$.

We have from earlier tasks that $L_U = E[U] = \frac{1}{\frac{\mu}{p \lambda}-1}$ and $L_N = E[N] = \frac{(1-p)\lambda \mu}{(\mu- \lambda)(\mu - p \lambda)}$.

Thus we have 

$$
1. W_U = \frac{L_U}{p \lambda} = \frac{1}{p \lambda(\frac{\mu}{p \lambda}-1)} = \frac{1}{\mu - p \lambda}
$$

and 

$$
2. W_N = \frac{L_N}{(1-p) \lambda} = \frac{(1-p)\lambda \mu}{(\mu- \lambda)(\mu - p \lambda)} \cdot \frac{1}{(1-p) \lambda} = \frac{\mu}{(\mu- \lambda)(\mu - p \lambda)}
$$

Which was what needed to be shown.

f)

Assume $\lambda = 5$ patients per hour and $\frac{1}{\mu} = 10$ minutes.

1. Plot $W_U$ and $W_N$ as functions of $p$.

```{r,echo=FALSE}
mu = 0.1*60    #Minutes to hours
lambda = 5


W_U <- function(p){
  return(1/(mu-p*lambda))
}

W_N <- function(p){
  return(mu/((mu-lambda)*(mu - p*lambda)))
}

P = seq(0, 1, length.out = 100)
WU = W_U(P)
WN = W_N(P)

plot(NULL, NULL, xlim=c(0, 1), ylim=c(min(c(WU, WN)) , max(c(WU, WN))), xlab="p", ylab="Expected waiting time")
  #Creates a figure and makes labels and titles.
  lines(P, WU, col="red")
  lines(P, WN, col="blue")
  legend(x="topleft", legend=c("Normal", "Urgent"), col=c("blue", "red"), lty=1:1)

```

2. Give interpretations of what the situations $p \approx 0$ and $p \approx 1$ describe.

The situation of $p \approx 0$ means that the probability of someone being an urgent patient is $\approx 0$ then there is only normal patients, not any urgent patients. 

And the situation of $p \approx 1$ means that the chance of someone being an urgent patient is $\approx 1$, there is only urgent patients, not any normal patients. 

3. Calculate the expected time spent at the UCC for a normal patient, $W_N$ in the extreme cases $p \approx 1$, $p \approx 0$.

For $p = 0, \lambda = 5, \mu = 6$ we have 

$$
W_N = \frac{6}{(6-5)\cdot (6-0)} = 1 
$$

The average time spent as a normal patient in the UCC is one hour for the case $p \approx 0.$

For $p = 1, \lambda = 5, \mu = 6$ we have 

$$
W_N = \frac{6}{(6-5)\cdot (6-1 \cdot 5)} = 6
$$

The average time spent as a normal patient in the UCC is six hours for the case $p \approx 1$.

3. We now want to calculate by hand the $p$ when expected time spent at the UCC for a normal patient is 2 hours. 

We get the following expression:

$$
2 = \frac{6}{6-5p} \implies 12- 10p = 6 \implies p = \frac{6}{10} = \frac{3}{5}
$$
Thus if the waiting time for a normal patient is 2 hours, then the chance of someone being a urgent patient is $p = \frac{3}{5}$.

g) NOT STARTED

__Problem 2 - Calibrating climate models__

a)

We want to start by finding the mean vector $\vec \mu$ and the conditional covariance matrix $\Sigma$.

Let $\vec X_A \sim N_{n_A} (\vec \mu_A, \Sigma_A)$ be $Y(\theta)$ with $n_A = 51$. 

Further let $\vec X_B \sim N_{n_B} (\vec \mu_B, \Sigma_B)$ be $Y(\theta)$ with $n_B = 5$. 

We thus have

$$
\vec X = (\vec X_A, \vec X_B) \sim N_{n_A + n_B}(\begin{bmatrix}
\vec \mu_A\\
\vec \mu_B
\end{bmatrix}, \begin{bmatrix}
\Sigma_{AA} & \Sigma_{AB}\\
\Sigma_{BA} & \Sigma_{BB}
\end{bmatrix})
$$
We have that $\vec \mu_A = \begin{bmatrix} 0.5, 0.5... 0.5\end{bmatrix}^T$ and $\vec \mu_B = \begin{bmatrix} 0.5, 0.5... 0.5\end{bmatrix}^T$ with lengths $n_A = 51, n_B = 5$.

We find the covariance matrix

$$
\begin{aligned}
Cov(Y(\theta_1),Y(\theta_2)) &= \sqrt{Var[Y(\theta_1)]Var[Y(\theta_2)]} \cdot Corr[Y(\theta_1),Y(\theta_2)]
\\
&= \sqrt{0.5^2 \cdot 0.5^2} \cdot (1 + 15 \cdot \mid \theta_1 - \theta_2 \mid) \cdot \exp(-15 \cdot \mid \theta_1 - \theta_2 \mid)
\\
&= 0.5^2\cdot (1 + 15 \cdot \mid \theta_1 - \theta_2 \mid) \cdot \exp(-15 \cdot \mid \theta_1 - \theta_2 \mid)
\end{aligned}
$$
Which gives us the following conditional distribution

$$
\vec X_C = (\vec X_A \mid \vec X_B = \vec x_B) \sim N_{n_A}(\vec \mu_C, \Sigma_C)
$$
For $\vec \mu_C = \vec \mu_A + \Sigma_{AB}\Sigma_{BB}^{-1}(\vec x_B - \vec \mu_B)$

and $\Sigma_C = \Sigma_{AA} -\Sigma_{AB}\Sigma_{BB}^{-1}\Sigma_{BA}$.

We now want to find a $90%$ prediction interval for $\mu_C$.


```{r,eval=TRUE}
theta = c(0.3,0.35,0.39,0.41,0.45)
y_theta = c(0.5,0.32,0.4, 0.35,0.6)

variance = 0.5*0.5


correlation <-function(a,b){
  return(1 + 15*abs(a-b)*exp(-15*abs(a-b)))
}


covariance <- function(a,b){
  #a,b = two different theta values
  n_a = length(a)
  n_b = length(b)
  
  cov_matrix <- matrix(n_a*n_b, n_a, n_b)
  
  for(i in 1:n_a){
    for(j in 1:n_b){
      cov_matrix[i,j] = correlation(a[i], b[j])*variance
    }
  }
  return(cov_matrix)
}

calculate_muSigma <- function(a,b,mu_a, mu_b, x_b){
  mu_c = mu_a + covariance(a,b) %*% solve(covariance(b,b)) %*% (x_b - mu_b)
  
  sigma_c = covariance(a,a) - covariance(a,b) %*% solve(covariance(b,b)) %*% covariance(b,a)
  
  return(list(mu_c,sigma_c))
}


a = seq(0.25,0.5, by = 0.005)
b = theta

mu_a = seq(0.5,0.5,length = 51)
mu_b = seq(0.5,0.5,length = 5)

b = theta
x_b = y_theta
  
res <- calculate_muSigma(a,b,mu_a,mu_b,x_b)

variance_c = diag(unlist(res[2]))
mu_c = unlist(res[1])


prediction_interval <- function(mu_c){
  U = mu_c
  L = mu_c
  n_c = length(mu_c)
  
  for(i in 1:n_c){
    p_i = 1.645*sqrt(variance_c[i])
    U[i] = U[i] + p_i
    L[i] = L[i] - p_i
  }
  
  return(list(U,L))
}

pred_int = prediction_interval(mu_c)


U = unlist(pred_int[1])
L = unlist(pred_int[2])


#plot(NULL, NULL, xlim=c(0.25, 0.5), ylim=c(min(L)*0.9, max(U)*1.1), main=expression(paste("90% Prediction interval")),
         #xlab=expression(paste("[", theta, "]")), ylab=expression(paste("y(", theta, ")")))
    #lines(a, L, col="green")
    #lines(a, U, col="blue")
    #lines(a, mu_c, col="red")
    #points(b, y_theta, col="black")
   # legend(x="topleft", legend=c("Lower bound of PI", "Upper bound of PI", expression(paste(mu["C"])), "Observed"), 
    #       col=c("green", "blue", "red", "black"), lty = c(1, 1, 1, NA), pch=c(NA, NA, NA,1))
```


